Water Quality at Toronto Beaches

### 2024-09-27 paper/paper.qmd
paper final changes

---
title: "Water Quality at Toronto Beaches"
subtitle: "An analysis of e-coli rates in toronto beaches 2007-2024"
author:
  - Jacob Gilbert
thanks: "Code and data are available at: https://github.com/JfpGilbert0/toronto-beaches-water-quality."
date: today
date-format: long
abstract: "This study analyses E. coli contamination trends at two of Toronto's popular beaches, Sunnyside Beach and Marie Curtis Park East Beach, from 2007 to 2024. Using data sourced from Open Data Toronto, the analysis examines temporal patterns in contamination levels, assessing the frequency and severity of unsafe E. coli concentrations as defined by public health thresholds. The study reveals that while E. coli levels appear less erratic and generally lower in magnitude in recent years, the frequency of tests exceeding unsafe thresholds has not significantly diminished, indicating persistent contamination risks. These findings underscore the need for continued monitoring and targeted interventions to protect public health. This research aims to inform beach management practices and guide future efforts to improve water quality at Toronto’s beaches, Aswell as provide insights for swimmers on the risks that they are exposed to."
format: pdf
jupyter: python3
bibliography: references.bib
---

```{python}
#| echo: false
# import dependencies
import jupyter
```

@@

\newpage
```{python}
#| echo: false
#| label: fig-0a
#| fig.pos: "H"
# Execute the external Python script
with open(f"../code/model_generation/summary_stats.py") as script_file:
    exec(script_file.read())
summary_stats.round(0)
```
**Figure 0a** Shows counts of unique variables and the date range of the data
```{python}
#| echo: false
#| label: fig-0b
#| fig.pos: "H"
# Execute the external Python script
with open(f"../code/model_generation/summary_ecoli.py") as script_file:
    exec(script_file.read())
ecoli_stats.round(0)
```
**Figure 0b** Shows summary statistics for e-coli levels over the whole dataset

## Measurement
E-coli is measured in MPN/100 ml a measurement that reflects the e coli particles in every 100 ml of lake water. For our data presentation log values are used to give a better visualisation of the data. To obtain an accurate representation of the beach as a whole 12 different locations here used to gather data across the 2 beaches as to remove bias that one testing location may result in. The result is that we obtain an unbiased representation of e coli levels at each beach as a whole over more than a decade. Some missing data we encounter however is that although we have the collection date we do not observe the time that data was collected. Missing this data may lead to some bias as the time of day could impact the baseline levels of e-coli in the water. What this form of measurement does allow us to do is look at the differences in water quality at a date level, Weekly section shows there is a good distribution of data from each day of the week.
There are some gaps in the data however, in the monthly data we only find results from the summer months. As during the summer the impact of water quality is at its highest this doesn't impact the usefulness of the data, however it does result in missing context of levels during the rest of the year.

@@

# Conclusion
This analysis of E. coli contamination levels at Sunnyside Beach and Marie Curtis Park East Beach reveals critical insights into the patterns and trends of water quality in Toronto’s public swimming areas. While the data shows a reduction in the variability of E. coli levels in recent years (assuming this chang in variability is not due to some bias), the frequency of unsafe readings remains concerning, particularly during peak summer months. The persistent risk highlighted by figure 2 indicates that ongoing contamination remains a public health challenge, despite efforts to improve water quality.Average tests that are above the target level is still over 30% in recent years. Policy makers should improve their efforts to reduce the number of days that water quality is unacceptable over the summer months.
The latter months are showing the greatest risk as shown in the monthly results section, thus efforts should be made to try and carry the lower levels shown earlier throughout the summer.
From the perspective of swimmers the results displayed here give insights on how to lower the risks of visiting the beaches when water quality is lower. Firstly overall we see that in recent years the extremes are much lower which is good news for lake swimmers in toronto. We also see that choosing to swim in the earlier months of the summer appears to bring with them better overall water quality than later months, May is especially low risk. Similarly the weekends bear a far lower risk to swimmers, which is fortunate as it is likely that these are the beaches busier times. The weekly data does have warning signs also, avoiding swimming during the week appears to be the consensus, especially at Sunnyside. The weekends, which are assumed to be busier for Toronto's beaches actually exhibit low risk, especially over recent years, this is a very positive sign for public health. If this is a result of policy then the data shows that it has been very successful in raising water quality on those days.
Overall the results here are fairly concerning when it comes to the percentage of results well above and acceptable threshold. There are silver linings however, and the insights gathered here can aid swimmers in choosing when and where they can minimise health risks from low water quality. The positive data points in the results show where efforts to improve water quality have been successful and where these efforts should be duplicated.

>>>>

---
title: "Water Quality at Toronto Beaches"
subtitle: "An analysis of e-coli rates in toronto beaches 2007-2024"
author:
  - Jacob Gilbert
thanks: "Code and data are available at: https://github.com/JfpGilbert0/toronto-beaches-water-quality."
date: today
date-format: long
abstract: "This study analyses E. coli contamination trends at two of Toronto's popular beaches, Sunnyside Beach and Marie Curtis Park East Beach, from 2007 to 2024. Using data sourced from Open Data Toronto, the analysis examines temporal patterns in contamination levels, assessing the frequency and severity of unsafe E. coli concentrations as defined by public health thresholds. The study reveals that while E. coli levels appear less erratic and generally lower in magnitude in recent years, the frequency of tests exceeding unsafe thresholds has not significantly diminished, indicating persistent contamination risks. These findings underscore the need for continued monitoring and targeted interventions to protect public health. This research aims to inform beach management practices and guide future efforts to improve water quality at Toronto’s beaches, Aswell as provide insights for swimmers on the risks that they are exposed to."
format: pdf
jupyter: python3
bibliography: references.bib
toc: true
toc-depth: 1
---
```{python}
#| echo: false
# import dependencies
import jupyter
```

@@

```{python}
#| echo: false
#| label: fig-0a
#| fig.pos: "H"
# Execute the external Python script
with open(f"../code/model_generation/summary_stats.py") as script_file:
    exec(script_file.read())
summary_stats.round(0)
```
**Figure 0a** Shows counts of unique variables and the date range of the data
```{python}
#| echo: false
#| label: fig-0b
#| fig.pos: "H"
# Execute the external Python script
with open(f"../code/model_generation/summary_ecoli.py") as script_file:
    exec(script_file.read())
ecoli_stats.round(0)
```
**Figure 0b** Shows summary statistics for e-coli levels over the whole dataset

## Measurement
E-coli is measured in MPN/100ml, a measurement that reflects the e coli particles in every 100 ml of lake water. For our data presentation log values are used to give a better visualisation of the data. To obtain an accurate representation of the beach as a whole 12 different locations here used to gather data across the 2 beaches as to remove bias that one testing location may result in. The result is that we obtain an unbiased representation of e coli levels at each beach as a whole over more than a decade. Some missing data we encounter however is that although we have the collection date we do not observe the time that data was collected. Missing this data may lead to some bias as the time of day could impact the baseline levels of e-coli in the water. What this form of measurement does allow us to do is look at the differences in water quality at a date level, Weekly section shows there is a good distribution of data from each day of the week.
There are some gaps in the data however, in the monthly data we only find results from the summer months. As during the summer the impact of water quality is at its highest this doesn't impact the usefulness of the data, however it does result in missing context of levels during the rest of the year.

@@

# Conclusion
This analysis of E. coli contamination levels at Sunnyside Beach and Marie Curtis Park East Beach reveals critical insights into the patterns and trends of water quality in Toronto’s public swimming areas. While the data shows a reduction in the variability of E. coli levels in recent years (assuming this change in variability is not due to some bias), the frequency of unsafe readings remains concerning, particularly during peak summer months. The persistent risk highlighted by figure 2 indicates that ongoing contamination remains a public health challenge, despite efforts to improve water quality.Average tests that are above the target level is still over 30% in recent years. Policy makers should improve their efforts to reduce the number of days that water quality is unacceptable over the summer months.
The latter months are showing the greatest risk as shown in the monthly results section, thus efforts should be made to try and carry the lower levels shown earlier throughout the summer.
From the perspective of swimmers the results displayed here give insights on how to lower the risks of visiting the beaches when water quality is lower. Firstly overall we see that in recent years the extremes are much lower which is good news for lake swimmers in toronto. We also see that choosing to swim in the earlier months of the summer appears to bring with them better overall water quality than later months, May is especially low risk. Similarly the weekends bear a far lower risk to swimmers, which is fortunate as it is likely that these are the beaches busier times. The weekly data does have warning signs also, avoiding swimming during the week appears to be the consensus, especially at Sunnyside. The weekends, which are assumed to be busier for Toronto's beaches actually exhibit low risk, especially over recent years, this is a very positive sign for public health. If this is a result of policy then the data shows that it has been very successful in raising water quality on those days.
Overall the results here are fairly concerning when it comes to the percentage of results well above and acceptable threshold. There are silver linings however, and the insights gathered here can aid swimmers in choosing when and where they can minimise health risks from low water quality. The positive data points in the results show where efforts to improve water quality have been successful and where these efforts should be duplicated.
